{fairmetrics}: An R package for group fairness evaluation

### 2025-06-01 paper.Rmd
Update paper.Rmd
current paper

---
title: '{fairmetrics}: An R package for group fairness evaluation'
tags:
  - R
  - Fairness
  - Machine Learning
  - Software
authors:
  - name: Benjamin Smith
    orcid: 0009-0007-2206-0177
    affiliation: 1
  - name: Jianhui Gao
    orcid: 0000-0003-0915-1473
    affiliation: 1
  - name: Jessica Gronsbell
    orcid: 0000-0002-5360-5869
    affiliation: 1
affiliations:
  - name: "Department of Statistcial Science, University of Toronto"
    index: 1
date: "2025-03-08"
bibliography: paper.bib
output:
  rticles::joss_article
  # md_document:
  #   preserve_yaml: TRUE
  #   variant: "markdown_strict"
journal: JOSS
---

@@

Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022]. Independence requires that the model’s classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups. Separation requires independence between the classifications and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within the positive (or negative) outcome class. Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model’s prediction is known, the protected attribute provides no additional information about the true outcome.  Below we summarize the fairness metrics that are available within the {fairmetrics} package.

@@

-   **(Positive) Predictive Parity:** Compares the positive predictive values across groups, assessing differences in the precision of positive predictions.

-   **(Negative) Predictive Parity:** Compares the negative predictive values across groups, assessing differences in the precision of negative predictions. [Note: Please add this to the package.]

@@

<!-- The figure looks good but do not use any abbreviations (i.e., Predicted probabilities).  Also, we use "protected" not sensitive attribute in paper so needs an update.  Also add negative predictive parity once added to the package. For combined: What's the difference between get_all and get_fairness? This isn't clear to me. Also I think you should have an argument for level for the CIs - that is, users input 0.95 for 95\% or 0.99 etc as that should be adjustable.   -->
The input to the {fairmetrics} package is a data frame or tibble containing the model's predicted probabilities, the true outcomes, and the protected attribute of interest. \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}. Users can evaluate a model for a specific criterion or multiple group fairness criteria using the combined metrics function..

![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow-v3.png)

@@

# Evaluate predictive equality
# (Setting message=FALSE to avoid cluttering the output)

get_fairness_metrics(
 data = test_data,
 outcome = "day_28_flg",
 group = "gender",
 group2 = "age",
 condition = ">=60",
 probs = "pred",
 cutoff = 0.41
)

#>                                       Metric GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
#> 1                         Statistical Parity        0.17      0.08       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
#> 2  Conditional Statistical Parity (age >=60)        0.34      0.21       0.13   [0.05, 0.21]  1.62 [1.18, 2.22]
#> 3                          Equal Opportunity        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
#> 4                        Predictive Equality        0.08      0.03       0.05   [0.02, 0.08]  2.67  [1.39, 5.1]
#> 5                 Balance for Positive Class        0.46      0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.41]
#> 6                 Balance for Negative Class        0.15      0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.74]
#> 7                          Predictive Parity        0.62      0.66      -0.04  [-0.21, 0.13]  0.94 [0.72, 1.23]
#> 8                         Brier Score Parity        0.09      0.08       0.01  [-0.01, 0.03]  1.12 [0.89, 1.43]
#> 9                    Overall Accuracy Parity        0.87      0.88      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
#> 10                        Treatment Equality        1.03      3.24      -2.21 [-4.35, -0.07]  0.32 [0.15, 0.68]
```

@@

```r
eval_eq_opp(
  data = test_data,
  outcome = "day_28_flg",
  group = "gender",
  probs = "pred",
  cutoff = 0.41
)

@@

For group-specific performance metrics, such as the True Positive Rate (TPR), False Positive Rate (FPR), Positive Predictive Value (PPV), Negative Predictive Value (NPV), and others, the `get_all_metrics()` function can be used. This function returns a data frame with the metrics calculated for each group.

[ NOTE: To me it is unclear why this is a separate function.  I think instead it makes sense to have an argument in your get_fairness_metrics function that turns the CI calculation on or off.  ]

```r
get_all_metrics(
  dat = test_data,
  outcome = "day_28_flg",
  group = "gender",
  probs = "pred",
  cutoff = 0.41
)

#>         Metric Group Female Group Male
#> 1           TPR         0.62       0.38
#> 2           FPR         0.08       0.03
#> 3           PPR         0.17       0.08
#> 4           PPV         0.62       0.66
#> 5           NPV         0.92       0.90
#> 6           ACC         0.87       0.88
#> 7   Brier Score         0.09       0.08
#> 8   FN/FP Ratio         1.03       3.24
#> 9 Avg Pred Prob         0.21       0.14
```

>>>>

---
title: '{fairmetrics}: An R package for group fairness evaluation'
tags:
  - R
  - Fairness
  - Machine Learning
  - Software
authors:
  - name: Benjamin Smith
    orcid: 0009-0007-2206-0177
    affiliation: 1
  - name: Jianhui Gao
    orcid: 0000-0003-0915-1473
    affiliation: 1
  - name: Jessica Gronsbell
    orcid: 0000-0002-5360-5869
    affiliation: 1
affiliations:
  - name: "Department of Statistcial Science, University of Toronto"
    index: 1
date: "2025-03-08"
bibliography: paper.bib
output:
  rticles::joss_article
  #rticles::arxiv_article
  # md_document:
  #   preserve_yaml: TRUE
  #   variant: "markdown_strict"
journal: JOSS
---

@@

Group fairness criteria are primarily classified into three main categories: independence, separation, and sufficiency [@barocas2023fairness; @Berk_Heidari_Jabbari_Kearns_Roth_2018; @Castelnovo_Crupi_Greco_Regoli_Penco_Cosentini_2022]. Independence requires that the model's classifications be statistically independent of the protected attribute, meaning the likelihood of receiving a positive prediction is the same across protected groups. Separation requires independence between the classifications and the protected attribute conditional on the true outcome, so that the probability of a positive prediction is equal across protected groups within the positive (or negative) outcome class. Sufficiency requires independence between the outcome and the protected attribute conditional on the prediction, implying that once the model's prediction is known, the protected attribute provides no additional information about the true outcome.  Below we summarize the fairness metrics that are available within the {fairmetrics} package.

@@

-   **Positive Predictive Parity:** Compares the positive predictive values across groups, assessing differences in the precision of positive predictions.

-   **Negative Predictive Parity:** Compares the negative predictive values across groups, assessing differences in the precision of negative predictions.

@@

<!-- The figure looks good but do not use any abbreviations (i.e., Predicted probabilities).  Also, we use "protected" not sensitive attribute in paper so needs an update.  Also add negative predictive parity once added to the package. For combined: What's the difference between get_all and get_fairness? This isn't clear to me. Also I think you should have an argument for level for the CIs - that is, users input 0.95 for 95\% or 0.99 etc as that should be adjustable.

BEN RESPONSE:

1. Updated the visual to not use abbreviations.
2. Changed "sensitive' attribute to "protected" attribute
3. Added negative predictive parity to package (`eval_neg_pred_parity()`; renamed `eval_pred_parity()` to `eval_pos_pred_parity()`). Submitted & Uploaded to CRAN. Added to the visual as well.
4.  Noted. Refactored the code to allow get_fairness_metrics() to cover the functionality of get_all_metrics(). This can be specified with the confint = FALSE.
5.  There is an argument level for the CIs, its the `alpha` parameter which by default is set to 0.05. Added the argument to the code. Do you want me to talk about it in the text as well? 
 -->
The input to the {fairmetrics} package is a data frame or tibble containing the model's predicted probabilities, the true outcomes, and the protected attribute of interest. \hyperref[workflow]{Figure ~\ref*{workflow}} shows the workflow for using {fairmetrics}. Users can evaluate a model for a specific criterion or multiple group fairness criteria using the combined metrics function.

![Workflow for using {fairmetrics} to evaluate model fairness across multiple criteria. \label{workflow}](fairmetrics-workflow.png)

@@

# Get fairness metrics
# Setting alpha=0.05 for 95% confidence intervals
get_fairness_metrics(
 data = test_data,
 outcome = "day_28_flg",
 group = "gender",
 group2 = "age",
 condition = ">=60",
 probs = "pred",
 cutoff = 0.41, 
 alpha = 0.05
)

#>                 Metric                          Full Metric Name GroupFemale GroupMale Difference    95% Diff CI Ratio 95% Ratio CI
#> 1                   PPR                        Statistical Parity        0.17      0.08       0.09   [0.05, 0.13]  2.12 [1.48, 3.05]
#> 2                   PPR Conditional Statistical Parity (age >=60)        0.34      0.21       0.13   [0.05, 0.21]  1.62 [1.18, 2.22]
#> 3                   FNR                         Equal Opportunity        0.38      0.62      -0.24 [-0.39, -0.09]  0.61 [0.44, 0.86]
#> 4                   FPR                       Predictive Equality        0.08      0.03       0.05   [0.02, 0.08]  2.67  [1.39, 5.1]
#> 5  Avg. Predicted Prob.                Balance for Positive Class        0.46      0.37       0.09   [0.04, 0.14]  1.24 [1.09, 1.41]
#> 6  Avg. Predicted Prob.                Balance for Negative Class        0.15      0.10       0.05   [0.03, 0.07]  1.50 [1.29, 1.74]
#> 7                   PPV                Positive Predictive Parity        0.62      0.66      -0.04  [-0.21, 0.13]  0.94 [0.72, 1.23]
#> 8                   NPV                Negative Predictive Parity        0.92      0.90       0.02  [-0.15, 0.19]  1.02 [0.79, 1.33]
#> 9           Brier Score                        Brier Score Parity        0.09      0.08       0.01  [-0.01, 0.03]  1.12 [0.88, 1.43]
#> 10             Accuracy                   Overall Accuracy Parity        0.87      0.88      -0.01  [-0.05, 0.03]  0.99 [0.94, 1.04]
#> 11          FN/FP Ratio                        Treatment Equality        1.03      3.24      -2.21   [-4.5, 0.08]  0.32 [0.15, 0.69]
```

@@

```r
eval_eq_opp(
  data = test_data,
  outcome = "day_28_flg",
  group = "gender",
  probs = "pred",
  confint = TRUE,
  cutoff = 0.41,
  alpha = 0.05
)

@@
